---
title: "Civic Density and American Political Behavior: How Residential Civic Diversity Shapes Political Participation "
date: today 
format:
  pdf: 
    word-count: "{{wordcount}}"
    number-sections: false
author:
  - name: Christian Hosam
    email: christian_hosam@berkeley.edu
    affiliation: 
      - name: Wellesley College
        city: Wellesley
        state: MA
        url: www.christianhosam.com
  - name: Norris Davis
    email: nldavis1@berkeley.edu
    affiliation: 
      - name: University of California, Berkeley
      - city: Berkeley
      - state: CA
      
keywords:
  - local politics, race, political geography
bibliography: recessions.bib
csl: american-political-science-association.csl
project: 
  execute-dir: project
---

```{r setup, include = FALSE}
# set global options
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = FALSE)
# Set the working directory to be the project directory (aka root or top-level directory)
knitr::opts_knit$set(root.dir = here:: here())
```

```{r Package Import, include=FALSE}
library(here)
library(knitr)
library(devtools)
library(tidyverse)
library(rticles)
library(bookdown)
library(kableExtra)
library(ggthemes)
library(stargazer)
library(pander)
```

## Introduction

Few concepts within the study of politics hold as much purchase as the critical nature of civic life in shaping citizen’s political engagement. When individuals are recruited, convened, and informed by civic organizations, participation is thought to follow. While this foundational truism is oft-repeated as a logical and necessary truth, historically it has been limited empirically by measures of the demand side of civic engagement: individual trust, networks, and dispositions, all of which have been in decline for some time \[CITE\]. Far less attention is placed on the *supply side* of civic opportunity; the civic infrastructure that may be available or present in a community, regardless of individual membership.

Part of the reason for this relative lack of scholarly attention is the lack of tools to do so. The demand side measures are largely drawn from self-reported survey measures that reflect individual's perception of their relationship to the state and their energy (or lack thereof) to participate in the democratic process. This is both substantively important and methodologically intuitive. But, as recent work on the nationalization of American politics, increased polarization, and the relationship between increased financial precarity and political decisionmaking highlights, people's individual choices are embedded within communities. Understanding the dynamics of local politics is thus important not only to evaluate demographic comparisons but also to understand what are the conditions under which people's decisions can be changed

## Civic Density: A New Tool for a Canonical Question

## Data and Methods

## Does Civic Density Predict Trump Support?

## Civic Density and Voter Turnout: A Youth Accelerant

## Civic Density and Partisan Identification

## Discussion

## Conclusion and Next Steps
